---
title: "Pikes Peak Race Data Analysis"
output: pdf_document
date: "2023-10-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document provides an analysis of the Pikes Peak race data for both male and female participants. We'll explore the data, clean it, and then generate visualizations to answer specific questions about the race results.

## Data Loading and Preprocessing

```{r load-data, message=FALSE, warning=FALSE}
#Load Libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(scales)

library(gt)
```

```{r}
#Load Data
females_data <- read.delim("/Users/mintraputlek/Library/CloudStorage/OneDrive-UniversityofMiami/UMIAMI/Classes/2023FALL/DIRPA Data Exercise 2023/DIRPA_Exer_PikesPeak_Females.txt", header=TRUE, stringsAsFactors=FALSE) %>%
  mutate(Gender = "Female")
males_data <- read.delim("/Users/mintraputlek/Library/CloudStorage/OneDrive-UniversityofMiami/UMIAMI/Classes/2023FALL/DIRPA Data Exercise 2023/DIRPA_Exer_PikesPeak_Males.txt", header=TRUE, stringsAsFactors=FALSE) %>%
  mutate(Gender = "Male")

combined_data <- bind_rows(females_data, males_data)
```


```{r}
#Format Time Function
format_time <- function(time_str) {
  parts <- unlist(strsplit(time_str, ":"))
  
  if (length(parts) == 1) {
    return(sprintf("00:00:%02d", as.integer(parts[1])))
  } else if (length(parts) == 2) {
    return(sprintf("00:%02d:%02d", as.integer(parts[1]), as.integer(parts[2])))
  } else if (length(parts) == 3) {
    return(sprintf("%02d:%02d:%02d", as.integer(parts[1]), as.integer(parts[2]), as.integer(parts[3])))
  } else {
    return(NA)
  }
}

format_hours_minutes_seconds <- function(x) {
  hours <- x %/% 3600
  x <- x %% 3600
  minutes <- x %/% 60
  seconds <- x %% 60
  
  sprintf("%02d:%02d:%02d", hours, minutes, seconds)
}
```

```{r}
#Data Cleaning
combined_data2 <- combined_data %>%
  mutate(
      Pace = gsub("[^0-9:]", "", Pace),
      Net.Tim = gsub("[^0-9:]", "", Net.Tim),
      Gun.Tim = gsub("[^0-9:]", "", Gun.Tim),
      Pace_hms = hms(sapply(Pace, format_time)),
      Net.Tim_hms = hms(sapply(Net.Tim, format_time)),
      Gun.Tim_hms = hms(sapply(Gun.Tim, format_time)),
      Pace_s = as.numeric(Pace_hms, "seconds"),
      Gun.Tim_s = as.numeric(Gun.Tim_hms, "seconds"),
      Net.Tim_s = as.numeric(Net.Tim_hms, "seconds"),
      AgeGroup = case_when(
        is.na(Ag) | Ag < 1 | Ag >= 70 ~"70+",   # Moved this condition to the top
        Ag >= 0 & Ag <= 14 ~ "0-14",
        Ag >= 15 & Ag <= 19 ~ "15-19",
        Ag >= 20 & Ag <= 29 ~ "20-29",
        Ag >= 30 & Ag <= 39 ~ "30-39",
        Ag >= 40 & Ag <= 49 ~ "40-49",
        Ag >= 50 & Ag <= 59 ~ "50-59",
        Ag >= 60 & Ag <= 69 ~ "60-69",
        TRUE ~ as.character(Ag)
    )
    ) %>%
  separate(Div.Tot, into = c("Division", "Total.Racers"), sep = "/") %>%
  mutate(
    Division = paste(Gender, AgeGroup, sep="-")
    ) %>%
  group_by(Division) %>%
  mutate(
    Division = cur_group_id()
    ) %>%
  ungroup()

```



## Question 1

What are the mean, median, mode, and range of the race results for all racers by gender?

```{r pressure, echo=FALSE}
#Q1 Boxplot
combined_data2 %>%
  ggplot(aes(x = Gender, y = Net.Tim_s)) +
  geom_boxplot(
            # custom boxes
        color="darkgreen",
        fill="darkgreen",
        alpha=0.2,
        
        # custom outliers
        outlier.colour="grey",
        outlier.fill="grey",
        outlier.size=3
  ) +
  stat_summary(fun=mean, geom="point", shape=4, color="black", size=3) +
  labs(title = "Distribution of Race Results by Gender",
       x = "Gender",
       y = "Race Result (in minutes)") +
 scale_y_continuous(labels = format_hours_minutes_seconds, breaks = seq(0, 14400, by=600)) +
  theme_minimal()


```

```{r}
# Table
combined_data2 %>%
  group_by(Gender) %>%
  summarise(
    Min_Time = min(Net.Tim_s, na.rm = TRUE),
    Max_Time = max(Net.Tim_s, na.rm = TRUE),
    Mean_Time = mean(Net.Tim_s, na.rm = TRUE),
    Median_Time = median(Net.Tim_s, na.rm = TRUE),
    Range_Time = Max_Time - Min_Time
  ) %>%
  mutate(
    Min_Time = seconds_to_period(Min_Time),
    Max_Time = seconds_to_period(Max_Time),
    Mean_Time = seconds_to_period(Mean_Time),
    Median_Time = seconds_to_period(Median_Time),
    Range_Time = seconds_to_period(Range_Time)
  )
```

```{r}
# First, generate the summarized data
summarized_data <- combined_data2 %>%
  group_by(Gender) %>%
  summarise(
    Min_Time = min(Net.Tim_s, na.rm = TRUE),
    Max_Time = max(Net.Tim_s, na.rm = TRUE),
    Mean_Time = mean(Net.Tim_s, na.rm = TRUE),
    Median_Time = median(Net.Tim_s, na.rm = TRUE),
    Range_Time = Max_Time - Min_Time
  )

```

```{r}
library(dplyr)

mode_by_gender <- combined_data2 %>%
  group_by(Gender) %>%
  count(Net.Tim_s) %>%
  arrange(Gender, desc(n)) %>%
  slice(1) %>%
  select(Gender, Net.Tim_s)

format_hours_minutes_seconds(2886)

format_hours_minutes_seconds(2128)



```

```{r}
#Histogram for Mode and Range
combined_data2 %>%
    ggplot(
           aes(x=Net.Tim_s, fill=Gender,text = Gender))+
      geom_histogram(
        alpha=0.7, color="black",  bins=33)+
      labs(x="Net Time (hh:mm:ss)", y="Number of Runners") +
      guides(fill=guide_legend(title="Gender"))+
      theme_minimal()+
      theme(
        axis.text.x = element_text(angle = 45,hjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        plot.title = element_text(size = 20),
        legend.text = element_text(size = 12)
      ) +
      scale_fill_manual(values=c("Male"="#DFE0E0", "Female"="#A7EE8A")) +  # Change colors here
      scale_x_continuous(labels = format_hours_minutes_seconds, breaks = seq(0, 14400, by=600)) +
      scale_y_continuous() 
        #ggtitle("Racer Time Distribution")
```
## Question 2

Analyze the difference between gun and net time race results.
```{r}
# Calculate the difference
combined_data2 <- combined_data2 %>%
  mutate(Difference = Gun.Tim_s - Net.Tim_s)

# Generate and print summary statistics
summary_stats <- combined_data2 %>%
  summarise(
    Mean_Difference = mean(Difference, na.rm = TRUE),
    Median_Difference = median(Difference, na.rm = TRUE),
    Std_Dev_Difference = sd(Difference, na.rm = TRUE)
  ) %>%   
  mutate(
    Mean_dif = seconds_to_period(Mean_Difference),
    Max_dif = seconds_to_period(Median_Difference),
    Std_Dev_dif = seconds_to_period(Std_Dev_Difference)
  )

print(summary_stats)

# Visualize the distribution
combined_data2 %>%
  group_by(
    Gender
  )%>%
  ggplot(aes(x = Difference)) +
  geom_histogram(fill = "#A7EE8A", color = "darkgreen", bins = 8) +
  labs(#title = "Distribution of Difference between Gun Time and Net Time",
       x = "Difference between Net and Gun Times (hh:mm:ss)",
       y = "Number of Racers") +
  scale_x_continuous(labels = format_hours_minutes_seconds, breaks = seq(0, 600, by=120)) +
  theme_minimal()

```


## Question 3

How much time separates Chris Doe from the top 10 percentile of racers of the same division?

```{r}

# Identify Chris Doe's division
chris_division <- combined_data2 %>%
  filter(Name == "Chris Doe") %>%
  pull(Division)

#Find Chris's Time 
chris_time <- combined_data2 %>%
  filter(Name == "Chris Doe") %>%
  pull(Net.Tim_s)

# Calculate the top 10 percentile time for that division
top_10_percentile_time <- combined_data2 %>%
  filter(Division == chris_division) %>%
  summarise(tenth_percentile = quantile(Net.Tim_s, 0.1)) %>%
  pull(tenth_percentile)

# Calculate Chris's result
chris_result <- round(chris_time - top_10_percentile_time)
chris_result_hms <- format_hours_minutes_seconds(chris_result)
chris_time_hms <- format_hours_minutes_seconds(chris_time)




# Filter the data for the division of interest
division_data <- combined_data2 %>%
  filter(Division == chris_division)
```


```{r}
division_data %>%
  ggplot(aes(x = "", y = Net.Tim_s)) + 
  geom_boxplot(fill = "gray", alpha = 0.5) +
  geom_jitter(data = subset(division_data, Name == "Chris Doe"), 
              aes(y = Net.Tim_s, color = "Chris Doe's Time"), 
              width = 0.2) + 
  geom_hline(aes(yintercept = top_10_percentile_time, color = "Top 10 Percentile"), 
             linetype = "dashed") +
  labs(#title = "Box Plot of Race Times",
       x = "",
       y = "Net Time (hh:mm:ss)") +
  scale_color_manual(name = "", 
                     values = c("Chris Doe's Time" = "lightgreen", 
                                "Top 10 Percentile" = "darkgreen")) +
  scale_y_continuous(labels = format_hours_minutes_seconds) +
  theme_minimal()

```

## Question 4


Compare the race results of each division.
```{r}
# Summarize the average race results for each division
division_summary <- combined_data2 %>%
  group_by(Division) %>%
  summarise(
    mean_time = mean(Net.Tim_s, na.rm = TRUE)
  )

# Create a column chart to compare average race results by division
race_results_plot <- ggplot(division_summary, aes(x = Division, y = mean_time)) +
  geom_col(aes(fill = Division)) +  # This creates the columns
  labs(
    title = "Comparison of Average Race Results by Division",
    x = "Division",
    y = "Average Time (seconds)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for better readability
        legend.position = "none")  # Hide legend as the fill colors correspond to the x-axis labels


  


```



```{r}
# Prepare data: Order data by Division and Net.Tim_s
combined_data2$Division <- factor(combined_data2$Division, levels = 1:16)

combined_data2 %>%
    ggplot(aes(x = Division, y = Net.Tim_s, fill = Gender)) +  # Add fill aesthetic here
    geom_boxplot(color = "darkgreen", alpha = 0.7) +
    labs(#title = "Box Plots of Race Times by Division", 
         y = "Time (seconds)", 
         x = "Division") +
    scale_fill_manual(values = c("grey", "lightgreen"), 
                      name = "Gender", 
                      breaks = c("Male", "Female")) +  # Set colors for male and female
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      scale_y_continuous(labels = format_hours_minutes_seconds, breaks = seq(0, 14400, by=600)) 



```





```{r}

gt_theme_espn <- function(data, ...){
  data %>% 
    opt_all_caps()  %>%
    opt_table_font(
      font = list(
        google_font("Lato"),
        default_fonts()
      )
    )  %>% 
    opt_row_striping() %>% 
    tab_options(
      row.striping.background_color = "#fafafa",
      table_body.hlines.color = "#f6f7f7",
      source_notes.font.size = 12,
      table.font.size = 16,
      table.width = px(700),
      heading.align = "left",
      heading.title.font.size = 24,
      table.border.top.color = "transparent",
      table.border.top.width = px(3),
      data_row.padding = px(7),
      ...
    ) 
}

winners_data <- combined_data2 %>%
  select(
    Division,
    Name,
    Hometown,
    Net.Tim_s
  ) %>%
  group_by(Division) %>%
  slice_min(Net.Tim_s, n = 1) %>%
  ungroup()

winners_data$FormattedTime <- seconds_to_period(winners_data$Net.Tim_s)

winners_data %>%
  gt() %>%
  tab_header(
    title = "Winners by Division"
  ) %>%
  fmt_number(columns = c("Net.Tim_s"), decimals = 2) %>%
  cols_hide(columns = "Net.Tim_s") %>%
  gt_theme_espn()



```




Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
